---
header-includes:
  - \usepackage{subfig}
  - \usepackage{float}
  - \floatplacement{figure}{H}
title: |
  | Asset Allocation and Investment Strategies
  | Assignment Three
author: |
  | Group 5
  | Business School, Imperial College London
  | CIDs: 02304617, 02325071, 02299241, 02280002, 02331035
date: "2023-02-22"
output: pdf_document
---

\newpage
\listoftables
\listoffigures
\newpage

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(dplyr)
library(data.table)
library(xtable)
library(knitr)

ff_25_valw <- read.csv("25_Portfolios_5x5.csv", skip = 15, nrows = length(16:1161))
colnames(ff_25_valw)[1] <- "month"
ff_25_valw$month <- as.Date(paste0(ff_25_valw$month, "01"), format = "%Y%m%d")
colnames(ff_25_valw)[-1] <- paste(colnames(ff_25_valw)[-1], "_valw", sep = "")

ff_25_eqw <- read.csv("25_Portfolios_5x5.csv", skip = 1165, nrows = length(1166:2311))
colnames(ff_25_eqw)[1] <- "month"
ff_25_eqw$month <- as.Date(paste0(ff_25_eqw$month, "01"), format = "%Y%m%d")
colnames(ff_25_eqw)[-1] <- paste(colnames(ff_25_eqw)[-1], "_eqw", sep = "")

ff_factors <- read.csv("F-F_Research_Data_Factors.csv", skip = 3, nrows = length(4:1149))
colnames(ff_factors)[1] <- "month"
ff_factors$month <- as.Date(paste0(ff_factors$month, "01"), format = "%Y%m%d")

ff_data <- merge(ff_25_valw, ff_25_eqw, by = "month")

ff_data <- merge(ff_factors, ff_data, by = "month") %>% as.data.table()

```

## Question 1(i)
We compute the excess return for the portfolios by subtracting the risk free rate from the mean return for the specified period and subsequently ran a regression for each portfolio, the results are shown below: 

```{r 1/1964 to 1/1993, echo = FALSE}

results_a <- data.frame()
results_b <- data.frame()
results_tval_a <- data.frame()
results_tval_b <- data.frame()
results_adj_r2 <- data.frame()

datatemp <- mom_ret_F3H6

datatemp <- merge(datatemp, fama5_data[, .(monthYear, mktrf, rf, smb, hml)], by = "monthYear", all.x = TRUE) %>% 
  select(top20, bottom20, rf, mktrf) %>% 
  as.data.table()

for (col in c(1, 2)) {
  model <- lm((datatemp[[col]] - datatemp$rf) ~ datatemp$mktrf)
  results_a <- rbind(results_a, coef(model)[1])
  results_b <- rbind(results_b, coef(model)[2])
  results_tval_a <- rbind(results_tval_a, coef(summary(model))[1, "t value"])
  results_tval_b <- rbind(results_tval_b, coef(summary(model))[2, "t value"])
  results_adj_r2 <- rbind(results_adj_r2, data.frame(adj_r2 = summary(model)$adj.r.squared))
} 


mean_ret <- matrix(colMeans(datatemp[,c(1,2)] - datatemp$rf)) #value-weighted
a <- matrix(results_a[,1]) #value-weighted
b <- matrix(results_b[,1]) #value-weighted
t_alphas <- matrix(results_tval_a[,1]) #value-weighted
t_betas <- matrix(results_tval_b[,1]) #value-weighted
adj_r2 <- matrix(results_adj_r2[,1]) #value-weighted

alphas <- data.frame(combo = paste(round(results_a[,1], 3), " (", round(results_tval_a[,1], 3), ")", sep = ""))
alphas <- matrix(alphas[,1])
betas <- data.frame(combo = paste(round(results_b[,1], 3), " (", round(results_tval_b[,1], 3), ")", sep = ""))
betas <- matrix(betas[,1])

alphas <- xtable(alphas)
betas <- xtable(betas)
mean_ret <- xtable(round(mean_ret, 3))
adj_r2 <- xtable(round(adj_r2, 3))
colnames(alphas) <- c("alphas")
rownames(alphas) <- c("Top20", "Bottom20")
colnames(betas) <- c("betas")
rownames(betas) <- c("Top20", "Bottom20")
colnames(mean_ret) <- c("meanreturn")
rownames(mean_ret) <- c("Top20", "Bottom20")
colnames(adj_r2) <- c("adjr2")
rownames(adj_r2) <- c("Top20", "Bottom20")

kable(mean_ret, align = "c", caption = "Portfolio Average Monthly Excess Returns")
kable(betas, align = "c", caption = "Portfolio Betas (t-values)")
kable(alphas, align = "c", caption = "Portfolio Alphas (t-values)")
kable(adj_r2, align = "c", caption = "Adjusted R-Squared")

```

## Question 1(ii)
Table 1, shows that the excess return increased towards the North-East direction indicating a value (high book-to-market value) and size (small market capitalization) premium on the portfolios.
It is also worth noting that growth portfolios (low book-to-market) do not exhibit a size premium. The excess returns decreases as the size reduces.

## Question 1(iii)
Table 2 shows that the $\betas$ are higher for portfolios with smaller market capitalization and lower book-to-market value. This implies that there should be a size and growth premium on the portfolios as indicated by CAPM. However Table 1 had  shown that there was indeed a size premium but also a value premium. The puzzle in the results is caused by the discrepancy between the actual excess return for the growth portfolios and the predicted excess return suggested by the CAPM. 

## Question 1(iv)
Table 3 shows the $\alphas$ of the regression and the corresponding t-values. Conducting a two-sided t-test with a 5% significance level, $t_{critical} = 1.96$. There are 12 significant portfolios, concentrated towards the north-east direction of the table (small market capitalization, high book-to-market value portfolios). It is also worth noting that $\alphas$ are not significant for any of the Big portfolios.
All significant portfolios have positive $\alphas$, implying that they are undervalued (in terms of returns) according to the CAPM.

The adjusted $R^2_{adj}$ shown in Table 4 are all moderately high (>60%), indicating the the regression equation (CAPM) can explain the excess returns of the portfolio sufficiently. However, the adjusted $R^2_{adj}$ reduces as we move towards the north-eastern direction (the same direction we saw for the $\alphas$ and excess returns)

## Question 1(v)
The plots of the predicted excess return versus actual excess return are shown below.

```{r 1/1964 to 1/1993 plots, echo = FALSE, fig.cap = "1964 - 1993: Predicted vs Actual Excess Return"}

rm(capm_plot)
capm_plot <- results_b*mean(ff_data_64_93$Mkt.RF)
capm_plot$act_ret <- colMeans(ff_data_64_93[,6:30] - ff_data_64_93$RF)
colnames(capm_plot)[1] <- "pred_ret"

plot(capm_plot$pred_ret, capm_plot$act_ret,
     pch = 16, col = "blue",
     main = "Predicted vs Actual Excess Returns",
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))
abline(a = 0, b = 1, col = "red")

```


## Question 1(vi)
Figure 1 shows that the CAPM fails to closely predict the excess returns of the portfolios during the period. Only one portfolio lies near the diagonal line represented by the CAPM. The CAPM model resulted in 8 (32%) portfolios having lower-than-predicted returns and 17 portfolios (68%) having higher-than-predicted returns. 

## Question 1(vii - viii)
We reproduced the plot from figure 1 in addition to categorizing the portfolios according to its respective size and B/M quintiles. The plots of the predicted excess return versus actual excess return are shown below.

```{r 1/1964 to 1/1993 group plots, echo = FALSE, , fig.cap = "1964 to 1993 Plots", fig.subcap = c("Predicted vs Actual Excess Returns (grouped by Size)", "Predicted vs Actual Excess Returns (grouped by B/M)"), out.width="50%"}

plot(capm_plot$pred_ret, capm_plot$act_ret,
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))
abline(a = 0, b = 1, col = "red")

group_size <- split(1:25, rep(1:5, each = 5))
symbols <- c(15, 17, 19, 4, 2)

for (i in 1:5) {
  points(capm_plot$pred_ret[group_size[[i]]], capm_plot$act_ret[group_size[[i]]], 
         pch = symbols[i], col = i)
  lines(capm_plot$pred_ret[group_size[[i]]], capm_plot$act_ret[group_size[[i]]], 
        col = i)
}

legend("topright", legend = c("Small", "ME2", "ME3", "ME4", "Big"), 
       pch = symbols, col = 1:5, bty = "n")

plot(capm_plot$pred_ret, capm_plot$act_ret,
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))
abline(a = 0, b = 1, col = "red")

group_bm <- list(list(1, 6, 11, 16, 21), 
                 list(2, 7, 12, 17, 22), 
                 list(3, 8, 13, 18, 23),
                 list(4, 9, 14, 19, 24),
                 list(5, 10, 15, 20, 25))
group_bm <- split(unlist(group_bm), rep(1:5, each = 5))
symbols <- c(15, 17, 19, 4, 2)


for (i in 1:5) {
  points(capm_plot$pred_ret[group_bm[[i]]], capm_plot$act_ret[group_bm[[i]]], 
         pch = symbols[i], col = i)
  lines(capm_plot$pred_ret[group_bm[[i]]], capm_plot$act_ret[group_bm[[i]]], 
        col = i)
}

legend("topright", legend = c("Low B/M", "BM2", "BM3", "BM4", "High B/M"), 
       pch = symbols, col = 1:5, bty = "n")

```

The CAPM suggests that growth portfolios should have higher risk and hence higher returns. From Figure 2(b), all five size groups of portfolios (from small market capitalization portfolios to large market capitalization portfolios), exhibit an inverse relationship as we move from value to growth, indicating the existence of a value premium which is inconsistent with the CAPM. 

Figure 2(a) shows that value portfolios (BM4 and High B/M) exhibit a positive relationship consistent with CAPM. As the Market Capitalization (Size) decreases, the return of the portfolio increases, implying the existence of a return premium for small firms. The growth portfolios (Low B/M, BM2, BM3) do not exhibit a positive linear relationship as indicated by CAPM, as the market capitalization (size) of the firm reduces, the excess return initially increases but eventually decreases (giving an 'n' shape)

Overall, the reliance of the CAPM model on systematic risk, $\betas$ and market risk premium might not be sufficient to predict excess returns in the stock market. As with our earlier analysis, the existence of a return premium in small-cap and high BM stocks shows the shortcomings of solely relying on the traditional CAPM model. Moreover, it also highlights the need for more sophisticated models that can capture the complexity of the financial markets which includes market frictions, other types of risk (idiosyncratic, liquidity, etc.) and macroeconomic uncertainties. 

**Value portfolios are represented by High Book-to-Market Value (High B/M), while Growth portfolios are represented by Low Book-to-Market Value (Low B/M).*

## Question 1(ix) - Part 1
We computed the predicted excess return for the period (1/1933 to 12/1963). The plots of the predicted excess return versus actual excess return are shown below.

```{r 1/1933 to 12/1963, echo = FALSE, fig.cap = "1933 - 1963: Predicted vs Actual Excess Return"}

ff_data_33_63 <- ff_data %>% 
  filter(month >= as.Date("1933-01-01") & month <= as.Date("1963-12-01")) %>% 
  as.data.table()

results_a <- data.frame()
results_b <- data.frame()
results_tval_a <- data.frame()
results_tval_b <- data.frame()
results_adj_r2 <- data.frame()

for (col in c(6:30)) {
  model <- lm((ff_data_33_63[[col]] - ff_data_33_63$RF) ~ ff_data_33_63$Mkt.RF)
  results_a <- rbind(results_a, coef(model)[1])
  results_b <- rbind(results_b, coef(model)[2])
  results_tval_a <- rbind(results_tval_a, coef(summary(model))[1, "t value"])
  results_tval_b <- rbind(results_tval_b, coef(summary(model))[2, "t value"])
  results_adj_r2 <- rbind(results_adj_r2, data.frame(adj_r2 = summary(model)$adj.r.squared))
} 

rm(capm_plot)
capm_plot <- results_b*mean(ff_data_33_63$Mkt.RF)
capm_plot$act_ret <- colMeans(ff_data_33_63[,6:30] - ff_data_33_63$RF)
colnames(capm_plot)[1] <- "pred_ret"

plot(capm_plot$pred_ret, capm_plot$act_ret,
     pch = 16, col = "blue",
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))
abline(a = 0, b = 1, col = "red")
```

Figure 3 shows that the CAPM also fails for the period (1/1933 to 12/1963), 92% of the portfolios are off the diagonal line represented by the CAPM. However, CAPM was able to better predict actual returns as compared to the previous time period of 1964 - 1993. 

```{r 1/1933 to 12/1963 group plots, echo = FALSE, fig.cap = "1933 to 1963 Plots", fig.subcap = c("Predicted vs Actual Excess Returns (grouped by Size)", "Predicted vs Actual Excess Returns (grouped by B/M)"), out.width="50%"}

plot(capm_plot$pred_ret, capm_plot$act_ret,
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))

abline(a = 0, b = 1, col = "red")

group_size <- split(1:25, rep(1:5, each = 5))
symbols <- c(15, 17, 19, 4, 2)

for (i in 1:5) {
  points(capm_plot$pred_ret[group_size[[i]]], capm_plot$act_ret[group_size[[i]]], 
         pch = symbols[i], col = i)
  lines(capm_plot$pred_ret[group_size[[i]]], capm_plot$act_ret[group_size[[i]]], 
        col = i)
}

legend("topright", legend = c("Small", "ME2", "ME3", "ME4", "Big"), 
       pch = symbols, col = 1:5, bty = "n")

plot(capm_plot$pred_ret, capm_plot$act_ret,
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))

abline(a = 0, b = 1, col = "red")

group_bm <- list(list(1, 6, 11, 16, 21), 
                 list(2, 7, 12, 17, 22), 
                 list(3, 8, 13, 18, 23),
                 list(4, 9, 14, 19, 24),
                 list(5, 10, 15, 20, 25))
group_bm <- split(unlist(group_bm), rep(1:5, each = 5))
symbols <- c(15, 17, 19, 4, 2)

for (i in 1:5) {
  points(capm_plot$pred_ret[group_bm[[i]]], capm_plot$act_ret[group_bm[[i]]], 
         pch = symbols[i], col = i)
  lines(capm_plot$pred_ret[group_bm[[i]]], capm_plot$act_ret[group_bm[[i]]], 
        col = i)
}

legend("topright", legend = c("Low", "BM2", "BM3", "BM4", "High"), 
       pch = symbols, col = 1:5, bty = "n")


```

The CAPM suggests that growth portfolios should have higher risk and hence higher returns. From Figure 4(a), only the small market capitalization portfolios (Small), do not exhibit a positive relationship as we move from value to growth; this is inconsistent with the CAPM. However, the larger market capitalization portfolios (ME2, ME3, ME4 and Big), exhibit a somewhat positive relationship as the book-to-market values of the portfolios increase, indicating high return for the level of risk; this is consistent with CAPM.

Figure 4(b) shows that value portfolios (BM3, BM4 and High B/M) exhibit a positive relationship consistent with CAPM. As the Market Capitalization (Size) reduces the return of the portfolio increases. Smaller firms are expected to have a higher level of risk and a higher return. The growth portfolios (Low B/M, BM2) do not exhibit a positive linear relationship as indicated by CAPM, as the market capitalization (size) of the firm reduces, the excess return initially increases but eventually decreases (giving an 'n' shape)

## Question 1(ix) - Part 2
We computed the predicted excess return for the period (1/1933 to 12/2021). The plots of the predicted excess return versus actual excess return are shown below.

```{r 1/1933 to 12/2021, echo=FALSE, fig.cap = "1933 - 1963: Predicted vs Actual Excess Return"}

ff_data_33_21 <- ff_data %>% 
  filter(month >= as.Date("1933-01-01") & month <= as.Date("2021-12-01")) %>% 
  as.data.table()

results_a <- data.frame()
results_b <- data.frame()
results_tval_a <- data.frame()
results_tval_b <- data.frame()
results_adj_r2 <- data.frame()

for (col in c(6:30)) {
  model <- lm((ff_data_33_21[[col]] - ff_data_33_21$RF) ~ ff_data_33_21$Mkt.RF)
  results_a <- rbind(results_a, coef(model)[1])
  results_b <- rbind(results_b, coef(model)[2])
  results_tval_a <- rbind(results_tval_a, coef(summary(model))[1, "t value"])
  results_tval_b <- rbind(results_tval_b, coef(summary(model))[2, "t value"])
  results_adj_r2 <- rbind(results_adj_r2, data.frame(adj_r2 = summary(model)$adj.r.squared))
} 


rm(capm_plot)
capm_plot <- results_b*mean(ff_data_33_21$Mkt.RF)
capm_plot$act_ret <- colMeans(ff_data_33_21[,6:30] - ff_data_33_21$RF)
colnames(capm_plot)[1] <- "pred_ret"

plot(capm_plot$pred_ret, capm_plot$act_ret,
     pch = 16, col = "blue",
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))
abline(a = 0, b = 1, col = "red")
```

Figure 5 shows that CAPM also fails for the period (1/1933 to 12/2021), 96% of the portfolios are off the diagonal line represented by the CAPM.

```{r 1/1933 to 12/2021 group plots, echo = FALSE, fig.cap = "1933 to 2021 Plots", fig.subcap = c("Predicted vs Actual Excess Returns (grouped by Size)", "Predicted vs Actual Excess Returns (grouped by B/M)"), out.width="50%"}
plot(capm_plot$pred_ret, capm_plot$act_ret,
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))

abline(a = 0, b = 1, col = "red")

group_size <- split(1:25, rep(1:5, each = 5))
symbols <- c(15, 17, 19, 4, 2)

for (i in 1:5) {
  points(capm_plot$pred_ret[group_size[[i]]], capm_plot$act_ret[group_size[[i]]], 
         pch = symbols[i], col = i)
  lines(capm_plot$pred_ret[group_size[[i]]], capm_plot$act_ret[group_size[[i]]], 
        col = i)
}

legend("topright", legend = c("Small", "ME2", "ME3", "ME4", "Big"), 
       pch = symbols, col = 1:5, bty = "n")

plot(capm_plot$pred_ret, capm_plot$act_ret,
     xlab = "Predicted Excess Return",
     ylab = "Actual Excess Return",
     xlim = c(min(capm_plot$pred_ret) - 0.1, max(capm_plot$pred_ret) + 0.1), 
     ylim = c(min(capm_plot$act_ret) - 0.1, max(capm_plot$act_ret) + 0.1))

abline(a = 0, b = 1, col = "red")

group_bm <- list(list(1, 6, 11, 16, 21), 
                 list(2, 7, 12, 17, 22), 
                 list(3, 8, 13, 18, 23),
                 list(4, 9, 14, 19, 24),
                 list(5, 10, 15, 20, 25))
group_bm <- split(unlist(group_bm), rep(1:5, each = 5))
symbols <- c(15, 17, 19, 4, 2)

for (i in 1:5) {
  points(capm_plot$pred_ret[group_bm[[i]]], capm_plot$act_ret[group_bm[[i]]], 
         pch = symbols[i], col = i)
  lines(capm_plot$pred_ret[group_bm[[i]]], capm_plot$act_ret[group_bm[[i]]], 
        col = i)
}

legend("topright", legend = c("Low", "BM2", "BM3", "BM4", "High"), 
       pch = symbols, col = 1:5, bty = "n")

```

The CAPM suggests that growth portfolios should have higher risk and hence higher returns. From Figure 6(b), the small market capitalization portfolios (Small, ME2), exhibit an inverse relationship as we move from value to growth inconsistent with the CAPM. While the medium and big market capitalization portfolios exhibit a positive relationship as the book-market value increases (as we move from value to growth portfolios) as implied by the CAPM.

Figure 6(a) shows that value portfolios (BM4 and High B/M) exhibit a positive relationship consistent with CAPM similar to the prior periods. As the Market Capitalization (Size) reduces the return of the portfolio increases. The growth portfolios (Low B/M, BM2, BM3) do not exhibit a positive linear relationship as indicated by CAPM, as the market capitalization (size) of the firm reduces, the excess return initially increases but eventually decreases (giving an 'n' shape).


